Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-04-13 manuscripts/TCS2023/tcs2023.tex
Fixed some overfull hboxes

It is worth noting that the recent method presented in~\cite{DBLP:conf/ictai/ChevalierFPZ19} for computing minimal trap spaces avoids the prime-implicant computation by relying on the \emph{most-permissive} semantics of Boolean networks.
This method has been implemented in the tool \texttt{mpbn}\footnote{\url{https://github.com/bnediction/mpbn}} demonstrated in~\cite{Paulev2020} for handling medium-sized models from the literature and very large synthetic models (up to 100,000 nodes).
However, this method is only applicable for \emph{locally-monotonic} Boolean networks, whereas the prime-implicants based method~\cite{klarner2015computing} is applicable for \emph{general} Boolean networks (i.e., including both locally-monotonic and non-locally-monotonic ones).
In addition, the \texttt{bioLQM} platform also provides another method using Binary Decision Diagrams (BDDs) in \url{http://colomoto.org/biolqm/doc/tools-trapspace.html}.
This method avoids the prime-implicant computation as it characterizes the set of generic trap spaces of a Boolean network by a BDD, then filters this set to get the set of all minimal trap spaces.
By this approach, it requires the computation of all solutions, whereas the methods~\cite{klarner2015computing,Paulev2020} based on Answer Set Programming (ASP) can start enumerating them as they are found.
Moreover, the main issue with the BDD-based method is that the number of generic trap spaces of a Boolean network may be extremely larger than its number of minimal trap spaces.
This issue limits the efficiency of the BDD-based method.
The study~\cite{DBLP:journals/tcs/NoualRS13} highlights the need for non-locally-monotonic Boolean networks in both biological and theoretical aspects.
Hence, it is still necessary to develop efficient methods for computing minimal trap spaces of large-scale general Boolean networks.

@@

This method was implemented using the state-of-the-art CP solver Chuffed\footnote{\url{https://github.com/chuffed/chuffed}}~\cite{demirovic2018solution} via its MiniZinc~\cite{nethercote2007minizinc} interface.
Because it is a high-level interface, the backtrack-and-replay method of~\cite{nabli2016enumerating} was not used but rather the alternative implementation with two global constraints for lexicographic ordering (ensuring enumeration of solutions) and iterated non-subset of each already found solution (for maximality).

@@

Fourth, regarding the comparison of the ASP-based methods (i.e., \texttt{PyBoolNet}, \texttt{mpbn}, and \texttt{Trappist-ASP}), we note that for all the models where \texttt{PyBoolNet} did not finish before the time limit, the timeout occurred during the computation of the prime-implicants.
Hence, not even a single minimal trap space was output by that method.
For all the remaining models, once \texttt{PyBoolNet} went through the prime-implicant phase, its ASP solving phase quickly returned the first 1000 minimal trap spaces, all under one second.
Hence, with the experimental results shown in this subsection as well as the two previous subsections, the practical differences between the ASP encoding of \texttt{Trappist-ASP} and that of \texttt{PyBoolNet} are not distinctly exposed.
The fact that our new ASP encoding is guaranteed to be linear in the number of nodes of the original model (see Subsection~\ref{subsec:computation_asp}) does not seem to be crucial here, however a much deeper analysis of those cases shall be shown in the next subsection.

@@

Regarding general Boolean networks, \texttt{Trappist-ASP} (even \texttt{Trappist-MaxSAT}) is far more efficient than both \texttt{bioLQM} and \texttt{PyBoolNet}.
The speedups of \texttt{Trappist-ASP} or \texttt{Trappist-MaxSAT} are large, even between one and three orders of magnitude for most models.
In addition, the experimental results also confirm that the ASP encoding of \texttt{Trappist-ASP} is much more efficient than that of \texttt{PyBoolNet}.

>>>>

It is worth noting that the recent method presented in~\cite{DBLP:conf/ictai/ChevalierFPZ19} for computing minimal trap spaces avoids the prime-implicant computation by relying on the \emph{most-permissive} semantics of Boolean networks.
This method has been implemented in the tool \texttt{mpbn}\footnote{\url{https://github.com/bnediction/mpbn}} demonstrated in~\cite{Paulev2020} for handling medium-sized models from the literature and very large synthetic models (up to 100,000 nodes).
However, this method is only applicable for \emph{locally-monotonic} Boolean networks, whereas the prime-implicants based method~\cite{klarner2015computing} is applicable for \emph{general} Boolean networks (i.e., including both locally-mono\-tonic and non-locally-monotonic ones).
In addition, the \texttt{bioLQM} platform also provides another method using Binary Decision Diagrams (BDDs) in \url{http://colomoto.org/biolqm/doc/tools-trapspace.html}.
This method avoids the prime-implicant computation as it characterizes the set of generic trap spaces of a Boolean network by a BDD, then filters this set to get the set of all minimal trap spaces.
By this approach, it requires the computation of all solutions, whereas the methods~\cite{klarner2015computing,Paulev2020} based on Answer Set Programming (ASP) can start enumerating them as they are found.
Moreover, the main issue with the BDD-based method is that the number of generic trap spaces of a Boolean network may be extremely larger than its number of minimal trap spaces.
This issue limits the efficiency of the BDD-based method.
The study~\cite{DBLP:journals/tcs/NoualRS13} highlights the need for non-locally-monotonic Boolean networks in both biological and theoretical aspects.
Hence, it is still necessary to develop efficient methods for computing minimal trap spaces of large-scale general Boolean networks.

@@

This method was implemented using the state-of-the-art CP solver Chuffed\footnote{\url{https://github.com/chuffed/chuffed}} \cite{demirovic2018solution} via its MiniZinc~\cite{nethercote2007minizinc} interface.
Because it is a high-level interface, the backtrack-and-replay method of~\cite{nabli2016enumerating} was not used but rather the alternative implementation with two global constraints for lexicographic ordering (ensuring enumeration of solutions) and iterated non-subset of each already found solution (for maximality).

@@

Fourth, regarding the comparison of the ASP-based methods (i.e.,\linebreak \texttt{PyBoolNet}, \texttt{mpbn}, and \texttt{Trappist-ASP}), we note that for all the models where \texttt{PyBoolNet} did not finish before the time limit, the timeout occurred during the computation of the prime-implicants.
Hence, not even a single minimal trap space was output by that method.
For all the remaining models, once \texttt{PyBoolNet} went through the prime-implicant phase, its ASP solving phase quickly returned the first 1000 minimal trap spaces, all under one second.
Hence, with the experimental results shown in this subsection as well as the two previous subsections, the practical differences between the ASP encoding of \texttt{Trappist-ASP} and that of \texttt{PyBoolNet} are not distinctly exposed.
The fact that our new ASP encoding is guaranteed to be linear in the number of nodes of the original model (see Subsection~\ref{subsec:computation_asp}) does not seem to be crucial here, however a much deeper analysis of those cases shall be shown in the next subsection.

@@

Regarding general Boolean networks, \texttt{Trappist-ASP} (even \texttt{Trappist\-MaxSAT}) is far more efficient than both \texttt{bioLQM} and \texttt{PyBoolNet}.
The speedups of \texttt{Trappist-ASP} or \texttt{Trappist-MaxSAT} are large, even between one and three orders of magnitude for most models.
In addition, the experimental results also confirm that the ASP encoding of \texttt{Trappist-ASP} is much more efficient than that of \texttt{PyBoolNet}.
